---
title: "Predict Exercise grades using Machine Learning"
output: html_document
---

## Reading the data
```{r Packages, echo=FALSE, results='hide', warning=FALSE,message=FALSE}
library(caret)
```

Initially, the training data is loaded into R. The read.csv function is used to load the training data into a data frame.

```{r Loading the data, echo=FALSE, cache=TRUE}
pml.data.train <- read.csv("pml-training.csv")
```

## Split into training and test set.
Using the caret, the training dataset is then split into a training and test dataset using the createDataPartition function. The data is randomly sampled with 60% of the data belonging to the training dataset. 

```{r split train, echo=FALSE, cache=TRUE}
inTrain <- createDataPartition(y=pml.data.train$classe, p=0.6, list=FALSE)
pml.train <- pml.data.train[inTrain,]
pml.test <- pml.data.train[-inTrain,]
```

## Cleaning the data

Looking at the columns in the training data set, it is noticible that many of the features have a large number of 'NA' fields. In order to build a model, all the features with large proportion of NA values are removed. Any feature more than 95% NA values are eliminated from the dataset. 

```{r remove NA, echo=TRUE, cache=TRUE}
temp <- apply(pml.train, 2, is.na)
temp <- apply(temp, 2, sum)
temp <- temp/nrow(pml.train)
non.na.cols <- temp < 0.95
pml.train.subset <- pml.train[,non.na.cols]
```

After removing the features with a high proportion of NAs, it is noticible that several of the remaining coloumns have a large number of blank entries. In order to check if these are useful, we use the nearZeroVar function to evaluate the variance of the features. A few of the features have near zero variance. These are also eliminated from the training data.

```{r nzv, echo=TRUE,cache=TRUE}
nzv <- nearZeroVar(pml.train.subset, saveMetrics = TRUE)
pml.train.subset <- pml.train.subset[,!nzv$nzv]
```

Finally, the first few columns of the dataset contain row numbers, user names and time related information. These features are not useful in building a predictive model and are eliminated from the datatset. 

```{r elminate time data, echo=TRUE, cache=TRUE}
non.useful.cols <- c(1:6)
pml.train.subset <- pml.train.subset[,-non.useful.cols]
```

As a result, we are left with a training dataset of `r ncol(pml.train.subset)` features. The same transformations are also applied to the testing subset of train dataset.

```{r test data, echo=TRUE, cache=TRUE}
pml.test.subset <- pml.test[,non.na.cols]
pml.test.subset <- pml.test.subset[,!nzv$nzv]
pml.test.subset <- pml.test.subset[,non.useful.cols]
```